---
title: "Group Project 1: Predictors of Quality of Life"
author: "Mahek Poddar,Kashyap Adipudi, Hoimo Chowdhury, Pratik Kulkarni"
subtitle: 'DATA5207: Data Analysis in the Social Sciences'
output:
  html_document:
    toc: no
    df_print: paged
  pdf_document:
    toc: TRUE
    toc_depth: 3
    latex_engine: xelatex
header-includes:
- \usepackage{caption}
- \captionsetup[figure]{font=scriptsize}
- \usepackage{sectsty}
- \allsectionsfont{\color{ForestGreen}}
mainfont: Avenir Book
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

\newpage

# Introduction

Quality of life is a multidimensional concept that captures an individual's subjective sense of well-being across physical, psychological, social, and material domains. Because it is a latent variable — one that cannot be directly observed or measured with a single question — it must be inferred from a set of related, observable indicators. This project uses the **2018 Australian World Values Survey (WVS)** to construct a composite measure of quality of life using factor analysis, and then examines a set of theoretically motivated predictors of that measure.

The WVS is a cross-national survey programme that has been running since the early 1980s. The 2018 Australian wave (n = 1,813) covers a wide range of values, beliefs, and attitudes — including social trust, personal values, and well-being indicators — making it well-suited for this investigation. These data were obtained from the [Australian Data Archive](https://dataverse.ada.edu.au/dataverse/wvs).

## More details on the group project

This assessable group project is completed over Lab 7 and Lab 8, and counts towards the group work component of DATA5207. In Lab 7 we create the dependent variable using factor analysis and develop our theory and descriptive analysis of predictors. In Lab 8 we will test our theories using regression models and assess the reliability of our scale.

\newpage

#TASK 1

**1. What are latent variables?**

Latent variables are constructs which can't be measured directly. Instead, their existence and magnitude can be determined through patterns of covariance from a set of observable (manifest) variables. Typically, any single question only captures a part of the construct and therefore can contain some measurement error. So researchers usually uses multiple indicators together to better approximate the underlying phenomenon. Some examples in social science are intelligence, personality traits, attitudes, and quality of life.

**2. What is factor analysis?**

Factor analysis is a statistical technique which is used to model the relationships of a set of observed variables in terms of a smaller number of unobserved common factors. The core idea is that the correlations between observed variables arise because they all reflect one or more underlying latent dimensions. This technique produces two key outputs, one is factor loading and the other one is factor scores. Factor loading express how strongly each observed variable is associated with each latent factor, and factor scores are individual-level estimates where each respondent falls on the latent dimension. In social sciences, factor analysis is used to create composite measures of constructs like well-being, attitudes, or personality.

**3. What are the assumptions of factor analysis?**

Factor analysis rests on several key assumptions. First, the variables included need to actually correlate with each other to some meaningful degree — if they are all independent there is nothing for a common factor to explain. Tools like the KMO measure and Bartlett's test can be used to check this before running the analysis. Second, the relationships between observed variables and the underlying factors are assumed to be linear, so non-linear patterns won't be captured well. Third, the variables should be roughly normally distributed, particularly when using maximum likelihood estimation. Fourth, the sample needs to be large enough relative to the number of variables — a common rule of thumb is around 5–10 respondents per variable with a minimum of about 150 cases overall. Finally, the variables cannot be perfectly correlated with each other (perfect multicollinearity), because this makes the correlation matrix impossible to invert and the analysis breaks down.

\newpage

# DLD

We start by loading in our libraries:

```{r, include=FALSE, message=FALSE, warning=FALSE, load-libraries}
library(ggplot2)
library(psych)
library(tidyverse)
library(ggpubr)
```

We load the 2018 Australian World Values Survey data:

```{r, include=FALSE, message=FALSE, warning=FALSE, load-data}
wvs.dat <- read.csv("wvs_data.csv")
```

\newpage

# Creating the Dependent Variable

Quality of life is a latent variable — we cannot directly measure it, but we can use several related survey items to construct a measure. We fit a factor analysis to five variables covering reported happiness, self-rated health, sense of freedom and control, life satisfaction, and financial satisfaction:

-   *Q46.* Taking all things together, would you say you are... (variable V10)
-   *Q47.* All in all, how would you describe your state of health these days? (variable V11)
-   *Q48.* How much freedom of choice and control do you feel you have over the way your life turns out? (variable V55)
-   *Q49.* All things considered, how satisfied are you with your life as a whole these days? (variable V23)
-   *Q50.* How satisfied are you with the financial situation of your household? (variable V59)

## Standardise Variables

We recode all the five variables so that the higher values indicate greater quality of life, and we remove missing values (coded as `-2` in the raw data):

```{r, echo = FALSE, message=FALSE, warning=FALSE,  recode-qol-vars}

wvs.dat <- wvs.dat %>%
  dplyr::mutate(
    
    # V10: reverse-coded so higher = happier (original: 1 = very happy, 4 = not at all happy)
    happiness = dplyr::recode(V10,
                              '1' = 4, '2' = 3, '3' = 2, '4' = 1,
                              '-2' = NULL),
    
    # V11: reverse-coded so higher = better health (original: 1 = very good, 5 = very poor)
    health = dplyr::recode(V11,
                           '1' = 5, '2' = 4, '3' = 3, '4' = 2, '5' = 1,
                           '-2' = NULL),
    
    # V59, V23, V55: already coded in positive direction; just remove -2
    finances = dplyr::recode(V59, '-2' = NULL, .default = V59),
    satisfaction = dplyr::recode(V23, '-2' = NULL, .default = V23),
    freedom = dplyr::recode(V55, '-2' = NULL, .default = V55)
    
  )

```

## Fitting the Factor Analysis

```{r,echo = FALSE, message=FALSE, warning=FALSE, factor-analysis}

fa.fit <- fa(wvs.dat[, c("happiness",
                          "health",
                          "finances",
                          "satisfaction",
                          "freedom")],
             nfactors = 1)

```

## Saving the Factor Scores

```{r, echo=FALSE, message=FALSE, warning=FALSE,  grab-factor-scores}

wvs.dat$life.quality <- as.numeric(fa.fit$scores)

head(wvs.dat$life.quality, 10)

```

\newpage

# Interpreting Factor Analysis

```{r, echo=FALSE, message=FALSE, warning=FALSE,  factor-analysis-output, echo=FALSE}

fa.fit 

```

The first column of the output table (`MR1`) shows how each item loads onto the latent quality of life factor. All loadings are positive, meaning that higher responses on each variable are associated with higher quality of life — which is the result we expect given our recoding. The `satisfaction` variable carries the highest loading (0.93), which makes sense: of the five items, self-reported life satisfaction is conceptually closest to what we are trying to measure.

The `h2` column shows the communalities — the proportion of each variable's variance explained by the factor. The `satisfaction` item again performs best (h2 = 0.86), while `health` has the lowest communality (h2 = 0.28), suggesting health is less tightly associated with the overall QoL factor than the other items. The `u2` column (uniqueness) is the inverse of h2, representing the proportion of each variable's variance *not* explained by the factor. Together, the factor accounts for approximately **49% of the total variance** across all five indicators.

## Visualising

```{r, echo = FALSE, message=FALSE, warning=FALSE,  fa-diagram, fig.cap = "Factor diagram showing loadings of each indicator onto the quality of life factor."}

fa.diagram(fa.fit)

```

## Distribution of Quality of Life

The QoL factor score is approximately standardised, with a mean of zero and a standard deviation of one. The distribution below shows this measure across the sample:

```{r, echo=FALSE, message=FALSE, warning=FALSE,  qol-histogram, fig.cap = "Distribution of quality of life measure produced by factor analysis.", fig.width = 5, fig.height = 4, fig.align = "center"}

ggplot(wvs.dat, aes(life.quality)) +
  geom_histogram(fill = 'black') +
  theme_minimal()

```

The distribution is approximately normal with a modest left skew, indicating that while most Australians report reasonably high quality of life, a smaller proportion report notably lower well-being.

\newpage

# Overview of Factor Analysis Steps

1.  Data cleaning and recoding
2.  Running the factor analysis
3.  Saving the factor scores
4.  Plotting the factor scores
5.  Interpreting factor analysis
6.  Analysing the association between quality of life and other variables (part of your assessable group work!)
7.  Assessing the reliability of your model (part of your assessable group work!)

\newpage

# Assessable Group Work

## Part 1 (Lab 7): Understanding Your Survey Data

### Step 1: Theory

We developed three theoretical propositions to guide our selection of predictor variables.

**Theory 1: Social trust predicts quality of life**

According to social capital theory trust between individuals and within communities is an important resource for an individuals well-being. Generally, individuals who trust others are more likely to engage into supportive social relationships. They are seen to feel secure in their environment, and shows low signs of psychological stress. We expect that respondents who express greater **general interpersonal trust** (V24: "Most people can be trusted") and greater **trust in their family members** (V102) will report higher quality of life.

*Predicted direction*: Higher trust → higher QoL.

**Theory 2: Valuing family and leisure predicts quality of life**

Research on subjective well-being and self-determination theory suggests that people whose values align with close relationships and restorative activities tend to experience greater well-being. Placing high importance on **family** (V4) and **leisure time** (V6) is likely to encourage investment in social bonds and opportunities for rest and recovery — both of which buffer against stress and promote positive affect.

*Predicted direction*: Higher importance placed on family and leisure → higher QoL.

**Theory 3: Sociodemographic resources predict quality of life**

Sociodemographic characteristics shape access to material resources, healthcare, social networks, and life opportunities — all of which influence well-being. We expect **education** (Q275) to be positively associated with QoL, as higher attainment is linked to better employment, health literacy, and social participation. We also expect **marital or partnership status** (V57) to predict QoL: a large body of empirical research finds that partnered individuals report higher well-being than those who are single, divorced, or widowed, owing to the social support, shared resources, and emotional stability that close intimate relationships provide.

*Predicted direction*: Higher education and being partnered → higher QoL.

------------------------------------------------------------------------

### Step 2: Selecting and Examining Predictors

Based on the theories above, we selected six predictor variables from the WVS. These are listed in the table below along with the relevant question wording from the survey.

| Variable | WVS Code | Survey Question |
|---|---|---|
| General interpersonal trust | V24 | Q57. Generally speaking, would you say that most people can be trusted, or that you need to be very careful in dealing with people? |
| Trust in family members | V102 | Q58. How much do you trust people in your family? |
| Importance of family | V4 | Q1. How important is family in your life? |
| Importance of leisure time | V6 | Q3. How important is leisure time in your life? |
| Highest education level | Q275 | Q275. What is the highest educational level that you have attained? |
| Marital/partnership status | V57 | Q273. Are you currently married, living together as married, divorced, separated, widowed, or single? |

#### Data Cleaning

```{r, echo=FALSE, message=FALSE, warning=FALSE,  recode-predictors}

wvs.dat <- wvs.dat %>%
  dplyr::mutate(
    
    # V24: binary trust variable; 1 = trusting, 0 = cautious
    trust_general = dplyr::recode(V24,
                                  '1' = 1L,
                                  '2' = 0L,
                                  '-2' = NULL),
    
    # V102: reverse-coded so higher = more trust in family
    trust_family = dplyr::recode(V102,
                                 '1' = 4L,
                                 '2' = 3L,
                                 '3' = 2L,
                                 '4' = 1L,
                                 '-2' = NULL),
    
    # V4: reverse-coded so higher = family more important
    imp_family = dplyr::recode(V4,
                               '1' = 4L,
                               '2' = 3L,
                               '3' = 2L,
                               '4' = 1L,
                               '-2' = NULL),
    
    # V6: reverse-coded so higher = leisure more important
    imp_leisure = dplyr::recode(V6,
                                '1' = 4L,
                                '2' = 3L,
                                '3' = 2L,
                                '4' = 1L,
                                '-2' = NULL),
    
    # Q275: ordinal education variable; higher = more education; -2 removed
    education = dplyr::recode(Q275,
                              '-2' = NULL,
                              .default = Q275),
    
    # V57: binary; 1 = partnered (married or cohabiting), 0 = not partnered
    partnered = dplyr::case_when(
      V57 %in% c(1, 2)       ~ 1L,
      V57 %in% c(3, 4, 5, 6) ~ 0L,
      TRUE                    ~ NA_integer_
    )
    
  )

```

#### Summary Statistics

```{r, echo = FALSE, message=FALSE, warning=FALSE, summary-stats}

wvs.dat %>%
  dplyr::select(trust_general, trust_family, imp_family,
                imp_leisure, education, partnered) %>%
  psych::describe() %>%
  dplyr::select(n, mean, sd, median, min, max, skew) %>%
  round(2)

```

The summary statistics reveal several features worth noting. Trust in family (mean = 3.76 out of 4) and importance of family (mean = 3.93 out of 4) both show ceiling effects, with most respondents selecting the highest or second-highest category. Education has a mean of approximately 5.7, corresponding roughly to post-secondary non-tertiary attainment, with meaningful spread across the range. Around 64% of respondents are partnered.

#### Distribution of Each Predictor

**General interpersonal trust (V24)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-trust-general, fig.cap = "Distribution of general interpersonal trust (V24).", fig.width = 5, fig.height = 3.5, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(trust_general)) %>%
  mutate(trust_label = ifelse(trust_general == 1,
                              "Most people can be trusted",
                              "Need to be very careful")) %>%
  ggplot(aes(x = trust_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q57. Dealing with people") +
  theme_minimal()

```

About 54% of respondents express general trust in most people. This is a binary variable with limited variance, but it is theoretically important as an indicator of broad social capital.

**Trust in family (V102)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-trust-family, fig.cap = "Distribution of trust in family members (V102, recoded: higher = more trust).", fig.width = 5, fig.height = 3.5, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(trust_family)) %>%
  mutate(trust_fam_label = factor(trust_family,
                                  levels = 1:4,
                                  labels = c("Do not trust at all",
                                             "Do not trust very much",
                                             "Trust somewhat",
                                             "Trust completely"))) %>%
  ggplot(aes(x = trust_fam_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q58. Trust in family (recoded)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

```

The vast majority of respondents trust their family completely or somewhat (78% and 19% respectively). The strong ceiling effect may limit this variable's ability to discriminate between QoL scores, though we retain it given its theoretical importance.

**Importance of family (V4)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-imp-family, fig.cap = "Distribution of importance of family (V4, recoded: higher = more important).", fig.width = 5, fig.height = 3.5, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(imp_family)) %>%
  mutate(imp_fam_label = factor(imp_family,
                                levels = 1:4,
                                labels = c("Not at all important",
                                           "Not very important",
                                           "Rather important",
                                           "Very important"))) %>%
  ggplot(aes(x = imp_fam_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q1. Importance of family (recoded)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

```

Over 90% of respondents rate family as "very important", producing a strong ceiling effect. We retain the variable for theoretical reasons but acknowledge this may reduce its predictive power.

**Importance of leisure time (V6)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-imp-leisure, fig.cap = "Distribution of importance of leisure time (V6, recoded: higher = more important).", fig.width = 5, fig.height = 3.5, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(imp_leisure)) %>%
  mutate(imp_leis_label = factor(imp_leisure,
                                 levels = 1:4,
                                 labels = c("Not at all important",
                                            "Not very important",
                                            "Rather important",
                                            "Very important"))) %>%
  ggplot(aes(x = imp_leis_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q3. Importance of leisure time (recoded)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 15, hjust = 1))

```

Leisure time is also rated highly, but with somewhat more variance than family importance: 41% rate it "very important" and 49% "rather important". This variable should discriminate better between QoL scores than the family importance item.

**Education level (Q275)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-education, fig.cap = "Distribution of education level (Q275).", fig.width = 6, fig.height = 4, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(education)) %>%
  mutate(edu_label = factor(education,
                            levels = c(1, 2, 3, 4, 5, 7, 8, 9),
                            labels = c("No formal",
                                       "Primary",
                                       "Lower secondary\n(Yr 9 or less)",
                                       "Upper secondary\n(Yr 10-12)",
                                       "Post-secondary\nnon-tertiary",
                                       "Bachelor",
                                       "Master",
                                       "Doctoral"))) %>%
  ggplot(aes(x = edu_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q275. Highest educational level attained") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

```

Education is spread across all categories, with the largest groups at post-secondary non-tertiary (24%), bachelor degree (30%), and upper secondary (21%). Category 6 does not appear in the Australian data — this is an artefact of the ISCED classification applied in the WVS. The variable is treated as ordinal.

**Marital / partnership status (V57)**

```{r, echo = FALSE, message=FALSE, warning=FALSE, plot-partnered, fig.cap = "Distribution of partnership status (V57, recoded to binary).", fig.width = 5, fig.height = 3.5, fig.align = "center"}

wvs.dat %>%
  filter(!is.na(partnered)) %>%
  mutate(part_label = ifelse(partnered == 1,
                             "Partnered\n(married or cohabiting)",
                             "Not partnered")) %>%
  ggplot(aes(x = part_label)) +
  geom_bar(fill = "grey30") +
  labs(x = NULL, y = "Count",
       title = "Q273. Marital status (recoded to binary)") +
  theme_minimal()

```

Around 64% of respondents are currently partnered. We recoded the original six-category variable (married, living together, divorced, separated, widowed, single) to a binary indicator, because our theoretical interest is in whether an individual has an intimate partner, not in the specific type of non-partnership situation.

------------------------------------------------------------------------

### Step 3: Written Sections

#### Introduction

This report investigates the predictors of quality of life among Australian adults using data from the 2018 World Values Survey.. Quality of life is an important but complex concept that includes material circumstances and also the psychology, social relationships and physical aspects of a person's well-being. Since quality of life is a latent variable, it cannot be directly measured but must be measured as a composite of five well-being measures through factor analysis. We then use established social science theories to create and test hypotheses regarding variables predicting high and low quality of life in Australia.

The knowledge of the factors that affect the quality of life has important repercussions for social policy. Specifically, by identifying which individual characteristics and social conditions are associated with positive well-being, interventions can then be developed based on these characteristics and the conditions to change people's lives for the better. The WVS provides a unique, well-rounded dataset, that contains information regarding user's attitudes, values, social connections, and demographic data collected from 1,813 nationally representative Australian adults.

#### Theory

We develop three theoretical accounts of what might predict quality of life.

*Social trust and quality of life.* Social capital theory argues that trust is one of the most fundamental resources for human well-being. When individuals trust their fellow citizens and their close relationships, they are able to engage more freely in cooperative behaviour, receive and provide social support, and maintain a sense of psychological security. We therefore expect that higher general interpersonal trust (V24) and greater trust in family members (V102) will be positively associated with quality of life.

*Personal values and quality of life.* Self-determination theory proposes that well-being is facilitated when individuals are able to invest in activities and relationships that are intrinsically meaningful to them. People who place high importance on family (V4) are likely to maintain stronger social support networks, and those who value leisure time (V6) are more likely to engage in restorative activities that protect against stress and burnout. We therefore expect that placing high importance on both family and leisure time will predict higher quality of life.

*Sociodemographic resources and quality of life.* Across a large body of empirical research, sociodemographic characteristics have been shown to shape well-being by determining access to economic resources, healthcare, and social opportunities. Education (Q275) is particularly important: more educated individuals tend to have better employment, higher incomes, greater health literacy, and more extensive social networks, all of which are associated with higher well-being. Partnership status (V57) is another robust predictor of well-being: partnered individuals report higher subjective well-being on average than those who are not partnered, owing to the social support, emotional intimacy, and economic benefits associated with close intimate relationships.

#### Data and Methods

This project uses the 2018 Australian World Values Survey, which covers 1,813 adults aged 18 and over recruited from across all Australian states and territories. The data were accessed through the Australian Data Archive. The WVS asks respondents about a wide range of topics including their values, social relationships, health, and well-being, which makes it a good fit for this kind of analysis.

To measure quality of life — our dependent variable — we used factor analysis on five survey items: happiness (V10), self-rated health (V11), sense of freedom and control over life (V55), overall life satisfaction (V23), and financial satisfaction (V59). Before running the factor analysis, we reverse-coded happiness and health so that a higher value always means better well-being, and we set the missing value code (`-2`) to `NA` across all five items. We specified a single-factor solution using the minimum residual method in the `psych` package. The resulting factor scores are approximately standardised, with a mean of zero and a standard deviation close to one — higher scores mean higher quality of life.

For our predictors, we made a few recoding decisions. General interpersonal trust (V24) was turned into a binary variable: respondents who said most people can be trusted were coded 1, and those who said you need to be careful were coded 0. Trust in family (V102), importance of family (V4), and importance of leisure time (V6) were each reverse-coded from their original scales so that higher values always mean more of the construct (e.g., more trust, more importance). Education (Q275) was kept as an ordinal variable running from 1 (no formal education) to 9 (doctoral degree) — note that category 6 does not appear in the Australian wave of the WVS, which is a feature of the international ISCED classification. Finally, marital status (V57) was simplified into a binary: 1 if the respondent is currently married or living with a partner, 0 otherwise. We made this choice because our theoretical interest is in whether someone has a close intimate partner, not in distinguishing among the different non-partnered situations. All remaining missing values were set to `NA`. In Lab 8, we will use linear regression to examine how these six predictors relate to the quality of life factor score.

\newpage

